---
title: Diversity of fish and vulnerable species in Marine World Heritage Sites based on OBIS data
date: "`r Sys.Date()`"
author: Pieter Provoost (edited by Chandra Earl)
output: (function(...) {
  rmdformats::robobook(toc_depth = 4, pandoc_args = c("+RTS", "-K2000m", "-RTS"), ...) })
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })  
---

This notebook explores species diversity in Marine World Heritage Sites using OBIS and GBIF data. This is work in progress.

## How to use this notebook

Optionally remove the following cache files before running the notebook:

- `data/occurrence.Rdata`: occurrences by site
- `data/bold.Rdata`: statistics on barcode sequences in BOLD for all fish species

If necessary, adjust fish and coral taxonomy below:

```{r message=FALSE, warning=FALSE}
fish_classes <- c("Actinopteri", "Cephalaspidomorphi", "Myxini", "Petromyzonti", "Elasmobranchii", "Holocephali", "Coelacanthi", "Chondrostei", "Teleostei")
fish_orders <- c("Cetomimiformes", "Gasterosteiformes", "Scorpaeniformes", "Stephanoberyciformes")
coral_classes <- c("Anthozoa")
not_coral_orders <- c("Penicillaria", "Spirularia", "Actiniaria")

fish_taxa_in_GBIF <- c("Coelacanthi","Dipneusti","Elasmobranchii","Holocephali","Myxini","Petromyzonti","Perciformes","Cypriniformes","Salmoniformes","Clupeiformes","Scorpaeniformes","Gadiformes","Pleuronectiformes","Anguilliformes","Tetraodontiformes","Siluriformes","Characiformes","Esociformes","Osmeriformes","Cyprinodontiformes","Gasterosteiformes","Atheriniformes","Beryciformes","Myctophiformes","Syngnathiformes","Aulopiformes","Mugiliformes","Beloniformes","Stomiiformes","Lophiiformes","Lepisosteiformes","Elopiformes","Ophidiiformes","Copelata","Zeiformes","Batrachoidiformes","Gymnotiformes","Osteoglossiformes","Percopsiformes","Gobiesociformes","Albuliformes","Stephanoberyciformes","Acipenseriformes","Amiiformes","Synbranchiformes","Notacanthiformes","Lampriformes","Gonorynchiformes","Saccopharyngiformes","Polypteriformes","Polymixiiformes","Cetomimiformes","Ateleopodiformes")
```

## Other resources

A separate repository has been set up for compiling species list based on existing publications, see https://github.com/iobis/mwhs-species-lists.

## Dependencies

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(caspr)
library(rredlist)
library(knitr)
library(ggplot2)
library(sf)
library(mapview)
library(concaveman)
library(sfheaders)
library(purrr)
library(robis)
library(stringr)
library(ftplottools)
library(ggrepel)
library(rgbif)
library(worrms)
```

## Compiling a list of all WoRMS species including IUCN Red List category and BOLD barcode statistics

First read all accepted WoRMS species. As it's not possible to get a full species list from the WoRMS or GBIF web services, I'm reading directly from an export provided by the WoRMS team. The list of fish classes and orders above is used to determine if species are fish or not. 

```{r message=FALSE, warning=FALSE}
worms <- read.csv("data/taxon.txt", sep = "\t", quote = "") %>%
  as_tibble() %>%
  filter(taxonRank == "Species" & taxonomicStatus == "accepted") %>%
  select(taxonID, scientificName, kingdom, phylum, class, order, family, genus) %>%
  distinct() %>%
  mutate(is_fish = class %in% fish_classes | order %in% fish_orders) %>%
  mutate(is_coral = class %in% coral_classes & !(order %in% not_coral_orders)) %>%
  mutate(aphiaID = as.integer(str_replace(taxonID, "urn:lsid:marinespecies.org:taxname:", ""))) %>%
  mutate_all(~na_if(., ""))
```

### Assign Red List categories

Here I'm using the `rredlist` package to get all Red List species from IUCN. Only the extinct and threatened categories are kept: `CR`, `EN`, `VU`, `EW`, `EX`.

```{r message=FALSE, warning=FALSE}
get_redlist_species <- function() {
  redlist <- tibble()
  page <- 0
  while (TRUE) {
    res <- rl_sp(page, key = "a936c4f78881e79a326e73c4f97f34a6e7d8f9f9e84342bff73c3ceda14992b9")$result
    if (length(res) == 0) {
      break
    }
    redlist <- bind_rows(redlist, res)
    page <- page + 1
  }
  redlist <- redlist %>%
    as_tibble() %>%
    filter(category %in% c("CR", "EN", "VU", "EW", "EX")) %>%
    mutate(category = factor(category, levels = c("EX", "EW", "CR", "EN", "VU"))) %>%
    group_by(scientific_name) %>%
    filter(row_number() == 1) %>%
    ungroup()
  return(redlist)  
}

redlist <- get_redlist_species()
```

Now we can label Red List species in the `worms` data frame.

```{r message=FALSE, warning=FALSE}
worms <- worms %>%
  left_join(redlist %>% select(scientific_name, category), by = c("scientificName" = "scientific_name"))
```

### Export WoRMS list

```{r message=FALSE, warning=FALSE}
openxlsx::write.xlsx(worms, "output/worms_species.xlsx", rowNames = FALSE)
write.csv(worms, "output/worms_species.csv", row.names = FALSE)
```

### Get BOLD barcode statistics for all fish species (SLOW)

Let's check BOLD for barcode sequences using all fish species in WoRMS:

```{r message=FALSE, warning=FALSE}
# fixes issue with bold package
# invisible(Sys.setlocale("LC_ALL", "C"))

if (!file.exists("data/bold.Rdata")) {
  fish_species <- worms %>% filter(is_fish) %>% pull(scientificName) %>% unique()
  bold_list <- sapply(fish_species, function(x) NULL)
  for (i in 1:length(bold_list)) {
    message(i, " ", names(bold_list)[i])
    if (is.null(bold_list[[i]])) {
      bold_list[[i]] <- tryCatch({
        caspr::bold_statistics(names(bold_list)[i])
      }, warning = function(warning_condition) {
      }, error = function(error_condition) {
      }, finally = {
      })
    }
  }
  save(bold_list, file = "data/bold.Rdata")
} else {
  load("data/bold.Rdata")
}

sequence_numbers <- unlist(map(bold_list, nrow))
fish_sequences <- tibble(species = names(sequence_numbers), sequences = sequence_numbers) %>%
  filter(sequences > 0)

worms <- worms %>%
  left_join(fish_sequences, by = c("scientificName" = "species"))
```

### Statistics

Now calculate some statistics:

```{r message=FALSE, warning=FALSE}
worms %>%
  summarize(
    species = n(),
    fish = sum(is_fish),
    vulnerable = length(na.omit(category)),
    vulnerable_fish = sum(is_fish * !is.na(category)),
    barcode_fish = sum(sequences > 0, na.rm = T)
  ) %>%
  kable(format.args = list(big.mark = ","))
```

Note that these numbers are slightly inflated due to homonyms, even after removing unaccepted names.

```{r message=FALSE, warning=FALSE}
ggplot(worms %>% filter(!is.na(category))) +
  geom_bar(aes(x = phylum, fill = category)) +
  coord_flip() +
  scale_fill_viridis_d(direction = -1)
```

## Marine World Heritage Sites statistics

In this section we will look at the diversity of fish and vulnerable species in each marine World Heritage site.

### Fetch spatial data

Spatial features for the marine World Heritage sites have been prepared in another repository, see <https://github.com/iobis/mwhs-shapes>.

```{r message=FALSE, warning=FALSE}
if (!file.exists("data/marine_world_heritage.gpkg")) {
  download.file("https://github.com/iobis/mwhs-shapes/blob/master/output/marine_world_heritage.gpkg?raw=true", "data/marine_world_heritage.gpkg")
}

shapes <- st_read("data/marine_world_heritage.gpkg")
```

### Process spatial data

For some sites, the GeoPackage has core as well as buffer areas. Merge the geometries by site.

```{r message=FALSE, warning=FALSE}
shapes_processed <- shapes %>%
  group_by(name) %>%
  summarize()
```

### Fetch occurrence data

#### OBIS

Now retrieve data from OBIS by area. OBIS is queried by bounding box, but a point-in-polygon calculation is used to discard points outside the areas.

```{r message=FALSE, warning=FALSE}
occ_for_geom <- function(geom) {
  wkt <- st_as_text(st_as_sfc(st_bbox(geom)), digits = 6)
  message(wkt)
  occ <- occurrence(geometry = wkt, fields = c("decimalLongitude", "decimalLatitude", "date_year", "scientificName", "aphiaID", "dataset_id")) %>%
    st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
  occ_filtered <- occ %>%
    filter(st_intersects(geometry, geom, sparse = FALSE)) %>%
    as_tibble() %>%
    select(-geometry)
  return(occ_filtered)
}

sf_use_s2(FALSE)

if (!file.exists("data/occurrence.Rdata")) {
  occs <- map(shapes_processed$geom, occ_for_geom)
  for (i in 1:nrow(shapes_processed)) {
    occs[[i]]$name <- shapes_processed$name[i]
  }
  occ <- bind_rows(occs)
  save(occ, file = "data/occurrence.Rdata")
} else {
  load("data/occurrence.Rdata")
}

occ <- occ %>%
  inner_join(worms %>% select(phylum, class, order, family, genus, is_fish, is_coral, aphiaID, category, sequences), by = "aphiaID") %>%
  mutate(
    aphia_fish = ifelse(is_fish == TRUE, aphiaID, NA),
    aphia_vulnerable = ifelse(!is.na(category), aphiaID, NA),
    aphia_vulnerable_fish = ifelse(is_fish == TRUE & !is.na(category), aphiaID, NA),
    aphia_barcode_fish = ifelse(is_fish == TRUE & sequences > 0, aphiaID, NA),
  ) 
```

#### GBIF

Now retrieve data from GBIF by area. GBIF is queried by bounding box, but a point-in-polygon calculation is used to discard points outside the areas. Since GBIF contains data other than marine occurrences, only fish are searched for here. AphiaIDs are also queried and recorded.

```{r message=FALSE, warning=FALSE}
occ_for_geom <- function(geom) {
  wkt <- st_as_text(st_as_sfc(st_bbox(geom)), digits = 6)
  message(wkt)
  occ_all <- NULL
  for (fishtaxa in fish_taxa_in_GBIF){
    #just get 10 first to check the number of records and to see if there is data
    occ_subset <- occ_data(geometry = wkt, scientificName = fishtaxa, limit = 10, start = 0)
    #if there is data
    if (!is.null(occ_subset$data)){
      #if there are more than 100000
      if (occ_subset$meta$count > 100000){
        #queries returning more than 100000 do not work since they are not allowed by GBIF
        message(fishtaxa)
        message(occ_subset$meta$count)
      } else {
        occ_subset <- occ_data(geometry = wkt, scientificName = fishtaxa, limit = 100000, start = 0)
        #filter columns
        occ_headers_filtered <- select(occ_subset$data, any_of(c("scientificName", "decimalLatitude", "decimalLongitude", "locality", "occurrenceID", "key", "year")))
        #add this fish taxa to the whole dataset
        occ_geog <- st_as_sf(occ_headers_filtered, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
        occ_filtered <- occ_geog %>%
            filter(st_intersects(geometry, geom, sparse = FALSE)) %>%
            as_tibble()
        occ_all <- bind_rows(occ_all, occ_filtered)
      }
    }
  }  
  return(occ_all)
}

sf_use_s2(FALSE)

if (!file.exists("data/GBIF_occurrence.Rdata")) {
  occs <- map(shapes_processed$geom, occ_for_geom)
  for (i in 1:nrow(shapes_processed)) {
    occs[[i]]$name <- shapes_processed$name[i]
  }
  occ <- bind_rows(occs)
  unique_sci_names <- name_parse(unique(occ$scientificName))[c("scientificname", "canonicalname")]

  names <- wm_name2id_(unique_sci_names$canonicalname)
  matching_names <- unique_sci_names$canonicalname %in% names(names)
  unique_sci_names$aphiaID[matching_names] <- names[unique_sci_names$canonicalname[matching_names]]
  occ <- merge(occ, unique_sci_names, by.x="scientificName", by.y="scientificname")
  
  save(occ, file = "data/GBIF_occurrence.Rdata")
} else {
  load("data/GBIF_occurrence.Rdata")
}

occ <- occ %>%
  inner_join(worms %>% select(phylum, class, order, family, genus, is_fish, is_coral, aphiaID, category, sequences), by = "aphiaID") %>%
  mutate(
    aphia_fish = ifelse(is_fish == TRUE, aphiaID, NA),
    aphia_vulnerable = ifelse(!is.na(category), aphiaID, NA),
    aphia_vulnerable_fish = ifelse(is_fish == TRUE & !is.na(category), aphiaID, NA),
    aphia_barcode_fish = ifelse(is_fish == TRUE & sequences > 0, aphiaID, NA),
  ) 
```

### Calculate statistics

```{r message=FALSE, warning=FALSE}
occ %>%
  summarize(
    records = n(),
    species = length(unique(aphiaID)),
    fish = length(unique(aphia_fish)),
    vulnerable = length(unique(aphia_vulnerable)),
    vulnerable_fish = length(unique(aphia_vulnerable_fish)),
    barcode_fish =  length(unique(aphia_barcode_fish))
  ) %>%
  kable(format.args = list(big.mark = ","))
```

```{r message=FALSE, warning=FALSE}
site_stats <- occ %>%
  group_by(name) %>%
  summarize(
    records = n(),
    species = length(unique(aphiaID)),
    fish = length(unique(aphia_fish)),
    vulnerable = length(unique(aphia_vulnerable)),
    vulnerable_fish = length(unique(aphia_vulnerable_fish)),
    barcode_fish =  length(unique(aphia_barcode_fish))
  )

xlsx::write.xlsx(site_stats, file = "output/sites.xlsx")

site_stats %>%
  kable(format.args = list(big.mark = ","))
```

```{r message=FALSE, warning=FALSE}
subset <- site_stats %>%
  filter(name %in% c("Great Barrier Reef", "Belize Barrier Reef Reserve System", "Ha Long Bay", "The Wadden Sea", "Heard and McDonald Islands", "Aldabra Atoll")) %>%
  arrange(name)

subset$x <- c(600, 100, 800, 500, 10, 60)
subset$y <- c(140, 80, 210, 3, 20, 40)

ggplot() +
  geom_segment(subset, mapping = aes(x = x, y = y, xend = fish, yend = vulnerable)) +
  geom_label(subset, mapping = aes(x = x, y = y, label = name), size = 3.5, label.size = 0) +
  geom_point(site_stats, mapping = aes(x = fish, y = vulnerable, size = species, color = species), shape = 21, stroke = 1.2, fill = "white") +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  scale_size(range = c(1, 8), breaks = c(1000, 5000, 10000)) +
  scale_color_viridis_c(end = 0.8, trans = "log10", breaks = c(1000, 5000, 10000)) +
  guides(color = guide_legend(), size = guide_legend()) +
  xlab("fish species") + ylab("vulnerable species") +
  ft_theme() +
  ggtitle("Fish and vulnerable species diversity at marine World Heritate sites")

ggsave("output/sites.png", width = 10, height = 6, dpi = 600)
```

### Species lists by site

```{r message=FALSE, warning=FALSE}
species_lists <- occ %>%
  group_by(name, scientificName, aphiaID, phylum, class, order, family, genus, is_fish, is_coral, category, sequences) %>%
  summarize() %>%
  select(site = name, scientificName, aphiaID, phylum, class, order, family, genus, is_fish, is_coral, iucn_category = category, bold_sequences = sequences) %>%
  arrange(site, phylum, class, order, family, scientificName) %>%
  ungroup()

write.csv(species_lists, file = "output/species_lists.csv", row.names = FALSE, na = "")
xlsx::write.xlsx(as.data.frame(species_lists), file = "output/species_lists.xlsx", showNA = FALSE, row.names = FALSE)
xlsx::write.xlsx(as.data.frame(species_lists %>% filter(is_fish)), file = "output/species_lists_fish.xlsx", showNA = FALSE, row.names = FALSE)
```
