---
title: Fish biodiversity in Marine World Heritage Sites
date: "`r Sys.Date()`"
author: Pieter Provoost
output: (function(...) {
  rmdformats::robobook(toc_depth = 4, pandoc_args = c("+RTS", "-K2000m", "-RTS"), ...) })
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })  
---

This notebook explores fish diversity in Marine World Heritage Sites using OBIS data.

## Fish species in WoRMS

[todo]

## Fish data in OBIS

```{r message=FALSE, warning=FALSE}
library(robis)
library(knitr)

if (!file.exists("checklist.Rdata")) {
  cl <- checklist(dropped = "include")
  cl <- cl %>%
    mutate(species_fish = ifelse(superclass == "Pisces", species, NA)) %>%
    mutate(species_vulnerable = ifelse(category %in% c("VU", "EN", "CR"), species, NA)) %>%
    mutate(species_vulnerable_fish = ifelse(superclass == "Pisces" & category %in% c("VU", "EN", "CR"), species, NA))
  save(cl, file = "checklist.Rdata")
} else {
  load("checklist.Rdata")
}

cl %>%
  summarize(
    records = sum(records),
    species = length(unique(species)),
    fish = length(unique(species_fish)),
    vulnerable = length(unique(species_vulnerable)),
    vulnerable_fish = length(unique(species_vulnerable_fish))
  ) %>%
  kable(format.args = list(big.mark = ","))
```

## Fish data for the Marine World Heritage Sites
### Fetch spatial data

First fetch the Marine World Heritage Sites shapefile from MarineRegions:

```{r message=FALSE, warning=FALSE}
library(sf)
library(dplyr)

if (!file.exists("shape/worldheritagemarineprogramme.shp")) {
  download.file("http://geo.vliz.be/geoserver/wfs?request=getfeature&service=wfs&version=1.0.0&typename=MarineRegions:worldheritagemarineprogramme&outputformat=SHAPE-ZIP", "shape.zip")
  unzip("shape.zip", exdir = "shape")
}

shapes <- st_read("shape/worldheritagemarineprogramme.shp") %>%
  select(full_name, country, geometry) %>%
  mutate(full_name = iconv(full_name, "latin1", "UTF-8"))
```

Let's create a map.

```{r message=FALSE, warning=FALSE}
library(mapview)
mapviewOptions(fgb = FALSE)

mapview(shapes)@map
```

### Process spatial data

```{r message=FALSE, warning=FALSE}
library(concaveman)
library(sfheaders)

shapes_processed <- shapes %>%
  mutate(geometry = sf_remove_holes(geometry)) %>%
  group_by(full_name) %>%
  summarize(geometry = st_union(geometry))

mapview(shapes, color = "red", lwd = 0.5, alpha = 1, alpha.regions = 0) +
  mapview(shapes_processed, color = "green", lwd = 0.5, alpha = 1, alpha.regions = 0, legend = FALSE)
```

### Fetch occurrence data

Here I'm fetching data from OBIS by area. OBIS is queried by bounding box, but a point-in-polygon calculation is used to discard points outside the areas.

```{r message=FALSE, warning=FALSE}
library(purrr)

occ_for_geom <- function(geom) {
  wkt <- st_as_text(st_as_sfc(st_bbox(geom)), digits = 6)
  message(wkt)
  occ <- occurrence(geometry = wkt, fields = c("scientificName", "species", "family", "order", "class", "superclass", "phylum", "decimalLongitude", "decimalLatitude", "date_year", "category", "dataset_id")) %>%
    st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
  occ_filtered <- occ %>%
    filter(st_intersects(geometry, geom, sparse = FALSE)) %>%
    as_tibble() %>%
    select(-geometry)
  return(occ_filtered)
}

if (!file.exists("occurrence.Rdata")) {
  occs <- map(shapes_processed$geometry, occ_for_geom)
  for (i in 1:nrow(shapes_processed)) {
    occs[[i]]$full_name <- shapes_processed$full_name[i]
  }
  occ <- bind_rows(occs)

  occ <- occ %>%
    mutate(species_fish = ifelse(superclass == "Pisces", species, NA)) %>%
    mutate(species_vulnerable = ifelse(category %in% c("VU", "EN", "CR"), species, NA)) %>%
    mutate(species_vulnerable_fish = ifelse(superclass == "Pisces" & category %in% c("VU", "EN", "CR"), species, NA))

  save(occ, file = "occurrence.Rdata")
} else {
  load("occurrence.Rdata")
}
```

### Calculate statistics

```{r message=FALSE, warning=FALSE}
occ %>%
  filter(!is.na(species)) %>%
  summarize(
    records = n(),
    species = length(unique(species)),
    fish = length(unique(species_fish)),
    vulnerable = length(unique(species_vulnerable)),
    vulnerable_fish = length(unique(species_vulnerable_fish))
  ) %>%
  kable(format.args = list(big.mark = ","))
```

```{r message=FALSE, warning=FALSE}
occ %>%
  filter(!is.na(species)) %>%
  group_by(full_name) %>%
  summarize(
    records = n(),
    species = length(unique(species)),
    fish = length(unique(species_fish)),
    vulnerables = length(unique(species_vulnerable)),
    vulnerable_fish = length(unique(species_vulnerable_fish))
  ) %>%
  kable(format.args = list(big.mark = ","))
```

## Fish DNA barcodes in molecular databases

[todo]
