---
title: Taxon IDs by World Heritage site
date: "`r Sys.Date()`"
author: Pieter Provoost
output: (function(...) {
  rmdformats::robobook(toc_depth = 4, pandoc_args = c("+RTS", "-K2000m", "-RTS"), ...) })
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })  
---

This simplified notebook generates lists of Aphia IDs by World Heritage site.

## How to use this notebook

Optionally remove the following cache files before running the notebook:

- `data/occurrence.Rdata`: occurrences by site

## Dependencies

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(sf)
library(purrr)
library(robis)
```

## Fetch spatial data

Spatial features for the marine World Heritage sites have been prepared in another repository, see https://github.com/iobis/mwhs-shapes.

```{r message=FALSE, warning=FALSE}
if (!file.exists("data/marine_world_heritage.gpkg")) {
  download.file("https://github.com/iobis/mwhs-shapes/blob/master/output/marine_world_heritage.gpkg?raw=true", "data/marine_world_heritage.gpkg")
}

shapes <- st_read("data/marine_world_heritage.gpkg")
```

## Process spatial data

For some sites, the GeoPackage has core as well as buffer areas. Merge the geometries by site.

```{r message=FALSE, warning=FALSE}
shapes_processed <- shapes %>%
  group_by(name) %>%
  summarize()
```

## Fetch occurrence data

Now retrieve data from OBIS by area. OBIS is queried by bounding box, but a point-in-polygon calculation is used to discard points outside the areas.

```{r message=FALSE, warning=FALSE}
occ_for_geom <- function(geom) {
  wkt <- st_as_text(st_as_sfc(st_bbox(geom)), digits = 6)
  message(wkt)
  occ <- occurrence(geometry = wkt, fields = c("decimalLongitude", "decimalLatitude", "date_year", "scientificName", "aphiaID", "dataset_id")) %>%
    st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
  occ_filtered <- occ %>%
    filter(st_intersects(geometry, geom, sparse = FALSE)) %>%
    as_tibble() %>%
    select(-geometry)
  return(occ_filtered)
}

sf_use_s2(FALSE)

if (!file.exists("data/occurrence.Rdata")) {
  occs <- map(shapes_processed$geom, occ_for_geom)
  for (i in 1:nrow(shapes_processed)) {
    occs[[i]]$name <- shapes_processed$name[i]
  }
  occ <- bind_rows(occs)
  save(occ, file = "data/occurrence.Rdata")
} else {
  load("data/occurrence.Rdata")
}
```

## Calculate statistics

```{r message=FALSE, warning=FALSE}
occ %>%
  summarize(
    records = n(),
    taxa = length(unique(aphiaID))
  )
```

```{r message=FALSE, warning=FALSE}
occ %>%
  group_by(name) %>%
  summarize(
    records = n(),
    taxa = length(unique(aphiaID))
  )
```

### Taxon lists by site

```{r message=FALSE, warning=FALSE}
taxon_lists <- occ %>%
  group_by(name, aphiaID) %>%
  summarize() %>%
  select(site = name, aphiaID) %>%
  arrange(site, aphiaID) %>%
  ungroup()

write.csv(taxon_lists, file = "output/taxon_lists.csv", row.names = FALSE, na = "")
```
